---
output:
  html_document:
    css: styles.css
    number_sections: yes
    toc: true
  word_document:
    reference_docx: QCPA_r-script-doc-styles.docx
    number_sections: yes
    toc: true
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
unlink(list.dirs()[endsWith(list.dirs(), "Output")], recursive=TRUE)
#rmarkdown::render("QCPA_r-script.Rmd", output_format = "all")
```
# Setup 
The following R script uses the Trichromatic example without UV (sea slugs seen by a triggerfish, *Rhinecanthus aculeatus*), but can similarly be applied to the Tetrachromatic example with UV (spiders seen by a bluetit, *Parus major*). To make the worked example readily accessible to users, we have prioritised using base R functions. As we will be working with primarily character type data (i.e., species names, region of interest descriptors), we will introduce the user to many base functions that are for list type data structures (e.g., *lapply()*). For clarity we recommend that, you as the user, need to be comfortable with the differences data types (e.g., numerical, character, factors) and data structures such as lists (collection of mixed data types) and data frames (special type of list). We also provide different methods to extract data, ideally introducing different coding methods as well as providing ways to increase efficiency of data wrangling.  

For the following R script to run, please ensure that you have implemented the QPCA batch script methods. In the following folder:

    Worked example data/Example1_tri-chromat_no_UV/Test Data

there should be two subfolders for the two species, *Aphelodoris varia* and *Aplysia sp*. Within these subdirectories, please confirm that data has been generated.


# Create the foundational dataset identifying unique animal IDs and regions of interest for multiple viewing distances

## Iterate over folders to extract species, animals and viewing distances

First, we will specify the location that the data files are located within with respect to where your working directory has been save to. Here we have assumed that the R script was saved in the parent folder of Worked example data. The next step will be to extract the names of the two focal species using the folder names within the *Test Data* folder. 

```{r Animal IDs}
pth<-getwd() #get directory; assumes the R script is saved in the parent folder of Worked example data
 
Extrct_Fold="Worked example data/Example1_tri-chromat_no_UV/Test Data" #set sub folder name containing your data

loc<-paste(pth,Extrct_Fold, sep="/")#Paste the two locations together to specify your global data location

Species=list.files(path=loc) #detects what species are in the data set. These could also be locations depending on your data set.

Species# should contain species folders and a log file

Species = Species[-length(Species)] #removes the log file

Species
```

Here we will demonstrate how to extract folder information names using a for loop. We first create an empty data frame called *Animal_Info*. Then, for each *i*$^{th}$ species (here *i* = 1 or 2), we create a path into the species folder, and record in the object *Indviduals_list* all the unique individual animal folder names using the *list.files* function. We then establish a second, nested loop, which will then iterate over each individual (*ind*) in *Indviduals_list*, to again extract all unique viewing distance folder names using *list.files*. To distinguish the viewing distance folders from other files saved within the directory, use the argument *pattern = "cm"* in *list.files*. Finally, we combine the unique species name, individual name and viewing distances into unique columns and add as a new row in *Animal_Info*.

```{r Animal Loop}
Animal_Info<-data.frame(NULL)# first create an empty data frame

# Then create a loop to iterate over the species folders, extracting viewing distances and individual animal IDs 
for (i in 1:length(Species)) {# here i will be either 1 or 2
    Ind_pth<-paste0(loc, "/",Species[i])# this pastes the ith species name from the Species object
    Indviduals_list<-list.files(path=Ind_pth)# extracts all individual folder names
    for (ind in 1:length(Indviduals_list)){# iterate over the number of individuals in each species folder
      # the variable dist lists all the unique viewing distance folder names identified by the "cm"
      Dists<-list.files(paste0(Ind_pth,"/", Indviduals_list[ind]),pattern = "cm")
      #we then add to Animal_Info a new row containing the species name, individual and viewing distance 
      rbind(Animal_Info, cbind(Species[i], Indviduals_list[ind], c(Dists)))-> Animal_Info
    }
}
colnames(Animal_Info)<-c("Species", "Ind", "Dist")# rename the column names in Animal_Info 
```

## Inspect data

Check that the data has correctly been imported and reflects your experimental records. We converted the character data into factors, which is the appropriate format for when we run statistical tests in downstream analyses. We use the *lapply* function to apply the function *factor* to each column in our data frame. We can again use the *lapply* function to apply the function *nlevels* to our newly converted data frame, which will report the number of unique elements within each column of the data frame *Animal_Info*.

```{r Check data}
#check data
Animal_Info
Animal_Info[,colnames(Animal_Info)] <-lapply(Animal_Info[,colnames(Animal_Info)],factor) # Convert character data into factors
str(Animal_Info)# observe structure of the data
lapply(Animal_Info, nlevels)# Check the number of unique species, individuals and viewing distances

# To visualize the experimental design we first aggregate the data to determine the number (length) of each individual by species and distance
Ani_info_tab<-aggregate(Ind~Species+Dist, data=Animal_Info, length)
barplot(Ind~Dist+Species, beside=TRUE, legend.text=TRUE, data=Ani_info_tab)
```
  
In this example there are only three individuals, however with a larger data set with missing data, this becomes an important step to determine a balanced experimental design. 

## Inspect regions of interest for inconsistent labelling (additional step)

The number of data files will depend on whether animal, background and/or animal+background options were selected; this will make up the prefix of the file name followed by an underscore and image analysis file type (e.g., '_*Summary Results.csv*'). For the following R code to work, regions of interest must be uniformly named and capitalised for all individuals analysed. Below is a method to ensure that these regions of interested have been correctly named by reporting all unique names across your data by examining all unique *.tif* file names. If you are confident that the naming was consistent, you may skip this step.  

```{r Check ROIs}
tmpAnidir<-apply(Animal_Info,1, function (x) paste(unlist(x), collapse = "/"))# create vector containing all the individual animal sub directories 
tmpAnidir <- paste(loc, tmpAnidir, sep="/") # paste the global file directories to the front of the sub directory paths
unique(unlist(strsplit(list.files(tmpAnidir, pattern=".tif"), "_"))[ c(TRUE,FALSE)])# This should highlight the unique regions of interest, if there are more than expected - look for naming inconsistencies
```
**How the nested line works.** *The above line can be broken down in four steps. 1. list.files() searches all the locations specified in the tmpAnidir vector for files with the ".tif" extension and puts these file names into a new character vector. 2. Knowing that file names are composed of two bits of information separated by an underscore, strsplit() cuts the filenames into two elements at the underscore and outputs a list containing these two new elements for each unique file. 3. unlist() puts all elements into a vector. As the region of interest are the prefix, it will be first element in the pair, which we can subset using: [ c(TRUE,FALSE)]. 4. Finally, the unique() function looks across all names used for each individual and retains only unique names, omitting duplicates.*

## Add regions of interest to Animal information dataset

To proceed with the next step, the regions of interest (ROIs) must be uniformly named across all individual specimens (see above if uncertain). Here, we introduce two methods to add ROIs. The first method is the quickest, and works if you are confident in your naming conventions. However, we provide a second method which extracts the ROIs from your data. 

### Simplified method to list ROIs

```{r simple ROIs}
ROI<-c("animal","animal+background","background")# add the names of ROIs
```

### Extract ROIs from the data

Here, we will utilise the batch script naming convention of using the ROI as the prefix of the image analysis file name followed by an underscore and image analysis file type (e.g., '_*Summary Results.csv*'). See Table 1 below for a list of all QCPA analyses and associated file names.  

```{r kable table exn out, echo=FALSE, warning=FALSE,comment=FALSE,error=FALSE, purl=FALSE}
library("kableExtra")
exns<- sort(unique(gsub("^[^_]+", "", list.files(
  paste0(loc,"/Aphelodoris varia/3_AphelodorisVaria_10_NF_D_RAW"),
  pattern=".csv", recursive = TRUE))))

exns<-as.data.frame(cbind(c(rep("Particle analysis",2), "GabRat","Particle analysis","LEIA","Particle analysis","CAA, VCA, BSA"), exns))
colnames(exns)<-c("Image analysis", "File name")
exns <- exns[order(exns$"Image analysis",exns$"File name"),]
rownames(exns)<-NULL
knitr::kable(exns, align="l",
             caption = "Table 1. QCPA image analysis and associated output files",
             format = "pipe", padding = 2)
```

We can then extract any particular image analysis .csv data files names using the *list.files* function specifying the file name as the pattern to search for. In the following code, we have used '_*Summary Results.csv*' as our string pattern. Once we have we have all the file names with '_*Summary Results.csv*', we remove this phrase using *gsub*.      

```{r ROI}
# Here we create a character vector determining whether animal, 
# background and/or animal+background options were used
# Using the first row of Animal_Info to specify the location to examine the number and types of _Summary Results.csv files 
ROI<- list.files(paste(loc,paste(unlist(Animal_Info[1,]), collapse = "/"),
                            sep="/"), pattern = "_Summary Results.csv")# find all the ROI output files for '_Summary Results.csv'
ROI<- gsub("_Summary Results.csv", "", ROI)# Here we remove the "_Summary Results.csv" information
ROI# In the QCPA batch script, we have investigated three ROIs
```

## Merge the list of ROIs names with the animal ID dataframe
```{r merge ROI}
Animal_Info_ROI<-merge(Animal_Info, ROI)# Expand each unique row of the Animal_Info data frame to included a unique ROI value
colnames(Animal_Info_ROI)[4]<-"ROI" #Name column 4 to ROI
Animal_Info_ROI$ROI<-factor(Animal_Info_ROI$ROI)
head(Animal_Info_ROI)# check data
str(Animal_Info_ROI)# check that ROI is a factor
```

# Create a function to import a csv list from VCA,BSA,CAA data, with options for particle analysis

As before, you could use a series of nested loops to extract the data in the csv files (e.g., section 2.1). However, here we introduce how to define a function to determine the data location using the data listed in the *Animal_Info_ROI* data frame and import the .csv file. By creating a function, we can then make our file importing dynamic, providing options to import different image analysis files. And, by using information called directly from the *Animal_Info_ROI* data frame, we ensure the data extracted is correct and reduces errors that would have potentially occurred if we did a broad import. First, we present a simple example of a function to import the VCA,BSA,CAA data files, and then, secondly, we expand on our function to demonstrate how to define and import multiple types of image analysis data files. 

## Simplified function example to read in image analysis files

In the following function, we walk through the steps for a function that only imports VCA,BSA,CAA image analysis data but will lay the foundations for when we create the function that can input any of the image analyses. There are three parts: (A) the first section will specify what type of image analysis we're interested in; (B) the second section extracts the data using *Animal_Info_ROI* data frame and stores the data as a list in the object *Sum_Res_list*; (C) finally, in the third section, we use the *do.call* function to extract all the list elements in *Sum_Res_list* and combine them into a data frame using *rbind*. From Table 1, the file suffix for this analysis is '_*Summary Results.csv*'. 
```{r reduced function}
# Part A.
# here we specify what image analysis so that we can use the correct file extension
exnfile="_Summary Results.csv"
Ani_ID_dat=Animal_Info_ROI #assign our animal ID data frame
path=loc # this is the global location set in section 2.1

# Part B.
# once the file extension condition is met, we use lapply to extract the data,
# using a nested function within lapply
Sum_Res_list<-lapply(1:nrow(Ani_ID_dat), # for each row in Ani_ID_dat...
                  function (x){ # start of our nested function
                    tmpLoc<- paste(unlist(Ani_ID_dat[x,]), collapse = "/")#extract the individual row information
                    tmpLoc<-paste0(paste(path,tmpLoc, sep="/"), exnfile)# create a directory string
                    data <- read.csv(tmpLoc)# read in the data using the newly specified location
                    data <- merge(Ani_ID_dat[x,], data)# add the specific animal information to the data
                    data #returns data
                  })
# Part C.
#we then collapse the Sum_Res_list list into a data frame by calling rbind in the function do.call
Sum_Res_df<-do.call(rbind, Sum_Res_list)

#check the imported data
dim(Sum_Res_df)# 18 rows by 133 columns
head(Sum_Res_df)[,1:12]# check the first six rows and first twelve columns, trimmed due to space


# Finally, to prevent any errors we remove the objects Ani_ID_dat, exnfile, path 
# from the global environment because we'll be recycling them in the second function
rm(exnfile); rm(Ani_ID_dat); rm(path)
```

## Dynamic function to read in various image analysis files
To dynamically import data, we are going to create the function *readQCPA*, which will take an animal ID data frame (*Ani_ID_dat*), the specific image analysis that we're interested in (*filetype*), and a folder directory (where we set the default *path* to the *loc* defined in section 1.2). For Part A of the function, we can conditionally set what image analysis we are interested in and set the *exnfile* object to the correct file extension. We have listed each image analysis and the argument name to be specified as the filetype in Table 2. As the last conditional (at the *else* level), we have used the *stop* function which will occur if *filetype* has not been specified, and will report an error. Part B is the same as in section 3.1 above, however it is worth noting that we used *merge* here instead of *rbind* (section 2.1) because this allows our function to handle data in both long or wide format. Then in Part C, we use the *return* function to output our newly imported data. 

```{r kable table ftyp, echo=FALSE, warning=FALSE,comment=FALSE,error=FALSE, purl=FALSE}
ftyp<-data.frame(Img_type=c("VCA,BSA,CAA", "Cluster Particle Analysis", "Cluster Results","Individual Particle"),
           ftyp=c('"VCA"','"PartAn"','"Clust"', '"IndParticle"') )
colnames(ftyp)<-c("Image analysis", "Filetype argument")
knitr::kable(ftyp, align="l",
             caption = "Table 2. QCPA image analysis and the different filetype arguments",
             format = "pipe", padding = 3)
```
  
  
```{r readQCPA function}
readQCPA<- function(Ani_ID_dat,filetype="NA",path=loc){
  #loc is assigned by default, but can be changed
  
  # Part A. 
  # here we specify arguments for different image analyses so that we can extract the correct file extension
  if(filetype=="VCA"){
    exnfile="_Summary Results.csv"
  } else if (filetype=="PartAn"){
    exnfile="_Cluster Particle Analysis Summary Results.csv"
  } else if (filetype=="Clust"){
    exnfile="_Cluster Results.csv"
  } else if (filetype=="IndParticle"){
    exnfile="_Individual Particle Results.csv"
  } else {
    stop('Please specify analysis output type: e.g., filetype =\n"VCA"\n"PartAn"\n"Clust"\n"IndParticle"')
  }# stop command halts the function and specifies how the user went wrong
  
  # Part B. 
  # once the file extension condition is met, we use lapply to extract the data,
  # using a nested function within lapply
  data_list<-lapply(1:nrow(Ani_ID_dat), function (x){# for each row in Ani_ID_dat...
    tmpLoc<- paste(unlist(Ani_ID_dat[x,]), collapse = "/")#extract the individual row information
    tmpLoc<-paste0(paste(path,tmpLoc, sep="/"), exnfile)# create a directory string
    data <- read.csv(tmpLoc)# read in the data using the newly specified location
    data <- merge(Ani_ID_dat[x,], data)# add the specific animal information to the data
    data #returns data
  })
  
  # Part C. 
  #we then collapse the data_list list into a data frame by calling rbind in the function do.call
  data_df<-do.call(rbind, data_list)
  return(data_df)#outputs the data frame
}
```

To demonstrate how the stop warning works, we have implemented the *readQCPA* function here without specifying the image analysis file type.  
```{r empty readQCPA, error=TRUE}
readQCPA(Animal_Info_ROI)# we include only the animal id dataset
```


We can then import our data using our function, and do some data cleaning by removing columns that are unnecessary.  

```{r implementing readQCPA}
# VCA,BSA,CAA analysis
VCA_analysis<-readQCPA(Animal_Info_ROI, filetype = "VCA") # read in and assign analysis 
#head(VCA_analysis)# inspect data
VCA_analysis<-VCA_analysis[, !colnames(VCA_analysis) %in% c("X", "Image")]# subset data excluding the column Image and X
dim(VCA_analysis)# 18 rows, by 131 columns
VCA_analysis[1:2,]#observe the first two rows of data

# Cluster Particle Analysis
Particle_analysis<-readQCPA(Animal_Info_ROI, filetype = "PartAn") # read in and assign analysis
#head(Particle_analysis)# inspect data
dim(Particle_analysis)# 52 rows, by 20 columns
colnames(Particle_analysis)[5]
colnames(Particle_analysis)[5]<-"ClusterID" # change column name to be consistent with Cluster analysis
Particle_analysis[1:2,]#observe the first two rows of data

# Cluster Results
Cluster_analysis<-readQCPA(Animal_Info_ROI, filetype = "Clust") # read in and assign analysis
#head(Cluster_Analysis)# inspect data
Cluster_analysis<-Cluster_analysis[, !colnames(Cluster_analysis) %in% c("X", "Image")]# subset data excluding the column Image and X
dim(Cluster_analysis)# 52 rows, by 21 columns
Cluster_analysis[1:2,]#observe the first two rows of data

# Individual Particle Analysis
IndParticle_analysis<-readQCPA(Animal_Info_ROI, filetype = "IndParticle") # read in and assign analysis
#head(Cluster_Analysis)# inspect data
IndParticle_analysis<-IndParticle_analysis[,colnames(IndParticle_analysis)!="X.1"]# subset data excluding the column X.1
dim(IndParticle_analysis)# 143 rows, by 23 columns
IndParticle_analysis[1:2,]#observe the first two rows of data
```

# Get a csv list from LEIA output

Here, we implement a function similar to section 3.1. However, because QCPA batch script exports LEIA analysis into a subdirectory, we have to add two additional folders to our location, *LEIA* and the ROI as a folder. For the LEIA folder, as we iterate over each row from *Animal_Info_ROI*, we convert the row into a character vector and add 'LEIA' as an element. To treat the ROI as a folder, we include a backslash before specifying the image analysis file extension, "_*Local Edge Intensity Analysis.csv*" (Table 1).

```{r LEIA CSV Data}
LEIA_Res_list<-lapply(1:nrow(Animal_Info_ROI), # for each row in Animal_Info_ROI
                  function (x){ # start of our nested function
                    tmpLoc <- as.character(unlist(Animal_Info_ROI[x,]))# convert row into a character vector
                    tmpLoc <- c(tmpLoc, "LEIA")[c(1:3,5,4)]# Add and order the new sub folder LEIA before ROI
                    tmpLoc <- paste(tmpLoc, collapse = "/")#paste tmploc into a directory character
                    tmpLoc <-paste0(paste(loc,tmpLoc, sep="/"), "/_Local Edge Intensity Analysis.csv")# create a directory string
                    data <- read.csv(tmpLoc)# read in the data using the newly specified location
                    data <- merge(Animal_Info_ROI[x,], data)# add the specific animal information to the data
                    data #returns data
                  })
#we then collapse the LEIA_Res_list list into a data frame by calling rbind in the function do.call
LEIA_Res_analysis<-do.call(rbind, LEIA_Res_list)
LEIA_Res_analysis<-LEIA_Res_analysis[, !colnames(LEIA_Res_analysis) %in% c("X", "Image")]# subset data excluding the column Image and X

#check the imported data
dim(LEIA_Res_analysis)# 18 rows by 37 columns
LEIA_Res_analysis[1:2,]#observe the first two rows of data
```

# Get a csv list from GabRat output

For GabRat analyses, there are no specific ROIs, so we only need the first three columns from *Animal_Info_ROI*. Similar to LEIA (section 4, above), QCPA batch script also exports GabRat analysis into a subdirectory. We will again iterate over each row of *Animal_Info_ROI* and convert the row into a character vector, adding 'GabRat' as an element. From Table 1, the GabRat file extension is: '_*GabRat_Results.csv*', where we place a backslash before the file name to specify that it is located within the GabRat folder. For this work example, we are only interested in the GabRat values '*dbl*', so we will then extract the rows that with only these values. We then extract the ROI information in a method similar to section 2.3.

```{r GabRat CSV Data}
# because we are not using ROI which adds three levels, we trim data frame using unique()
GabRat_Res_list<-lapply(1:nrow(unique(Animal_Info_ROI[,1:3])), # for each the 6 rows in Animal_Info_ROI
                  function (x){ # start of our nested function
                    tmpLoc <- as.character(unlist(Animal_Info_ROI[x,1:3]))# using only the first three columns, convert row into a character vector
                    tmpLoc <- c(tmpLoc, "GabRat")# Add the new sub folder GabRat
                    tmpLoc <- paste(tmpLoc, collapse = "/")# paste tmploc into a directory character
                    tmpLoc <-paste0(paste(loc,tmpLoc, sep="/"), "/_GabRat_Results.csv")# create a directory string
                    data <- read.csv(tmpLoc)# read in the data using the newly specified location
                    data <- merge(Animal_Info_ROI[x,1:3], data)# add the specific animal information to the data,using only the first three columns
                    data #returns data
                  })
#we then collapse the LEIA_Res_list list into a data frame by calling rbind in the function do.call
GabRat_Res_analysis<-do.call(rbind, GabRat_Res_list)

#check the imported data
dim(GabRat_Res_analysis)# 72 rows by 7 columns
head(GabRat_Res_analysis)# We only want every ~ fourth row for the dbl values

# set up a true/false condition on whether row has the pattern "_dbl"
GabRat_Res_analysis<-GabRat_Res_analysis[endsWith(GabRat_Res_analysis$ID, "_dbl"),]
rownames(GabRat_Res_analysis)<-NULL # reset row numbers

# Then we want to determine ROI where we split the ID column data on the shared pattern "WholeImage_"
# We then only take the second elemnt, using a FALSE,TRUE index 
GabRat_Res_analysis$ROI<-unlist(strsplit(GabRat_Res_analysis$ID, "WholeImage_"))[c(FALSE,TRUE)]
# then remove the '_dbl'
GabRat_Res_analysis$ROI<-gsub("_dbl","", GabRat_Res_analysis$ROI)

# remove unnecessary columns 
GabRat_Res_analysis<-GabRat_Res_analysis[, !colnames(GabRat_Res_analysis) %in% c("X","ID", "Sigma")]# subset data excluding the column X

#check the imported data
dim(GabRat_Res_analysis)# 18 rows by 5 columns
GabRat_Res_analysis[1:2,]#observe the first two rows of data
```

# Combine data output into csv lists across analyses and output data

## Merging data sets

To merge files, we will again use the *merge* function.

### Particle analyses

```{r merge two particle analyses files}
# Merging two files together
Cluster_Particle_analysis<-merge(Cluster_analysis, Particle_analysis, by=c("Species","Ind","Dist","ROI", "ClusterID"))
dim(Cluster_Particle_analysis)# 52 rows and 36 columns
Cluster_Particle_analysis[1:2,]#inspect data
```


### VCA,BSA,CAA data, LEIA output and GabRat analyses 
Here we demonstrate how to collate data across the VCA,BSA,CAA data, LEIA output and GabRat output. We have excluded the particle analysis outcome from this step.

```{r combine VCA LEIA and GabRat}
# Merge the first two files together
VCA_LEIA_GabRat_analysis<-merge(VCA_analysis, LEIA_Res_analysis, by=c("Species","Ind","Dist","ROI"))
VCA_LEIA_GabRat_analysis[1:2,]# inspect data
dim(VCA_LEIA_GabRat_analysis)# 18 rows and 164 columns

VCA_LEIA_GabRat_analysis<-merge(VCA_LEIA_GabRat_analysis, GabRat_Res_analysis, by=c("Species","Ind","Dist","ROI"))
VCA_LEIA_GabRat_analysis[1:2,]#inspect data again
dim(VCA_LEIA_GabRat_analysis)# 18 rows and 165 columns
```


## Output files
Depending on what image analysis sets you are using, we have provided multiple ways to save the image analysis data files as csvs in the folder 'Output'. First, we use *dir.create* to create the folder 'Output' in the Test Data folder, and save this new folder location as *Out_path*. To save a single file, you can use *write.csv*, where we have used the argument *row.names = FALSE* to make sure that the row numbers in the dataframe aren't included in the csv file. To save mutliple files, we utilise the fact that our naming convention in this R tutorial specifies image analysis data files with _*analysis*. Using the *ls* function with the pattern argument set to "_*analysis*", we can create a list of all our image analysis dataframes. Then using a for loop, we iterate over each dataframe, rename it using the current date, and then output the csv file. Finally, to check that the files have been created, we use *list.files* set to our Output folder location. 

```{r Output files}
dir.create(paste(c(loc, "Output"), collapse = "/"))# Create the Output directory
Out_path = paste(c(loc, "Output"), collapse = "/")

# For a single file
write.csv(GabRat_Res_analysis, row.names = FALSE, #remove row names
          file=paste(c(Out_path, "GabRat_Res_analysis.csv"), collapse = "/"))# save to the folder Output in Test Data

# For multiple files
savefile_list<-ls(pattern = "analysis")
savefile_list
for (i in savefile_list){
  # create file name and add the .csv extension
  # add date of creation using format(Sys.time(), "_%d_%b_%Y"), formatted to day_month_year
  fname=paste0(gsub("_analysis", "", i), "_data",format(Sys.time(), "_%d_%b_%Y"),".csv")# remove "_analysis" and add "_data" and date created
  write.csv(GabRat_Res_analysis, row.names = FALSE, #remove row names
          file=paste(c(Out_path, fname), collapse = "/"))# save to the folder Output in Test Data
}

# check that your files have saved
list.files(Out_path)
```


# Fast import for large data sets using data.table (optional - advanced)

Maybe not...? Provide a single demonstration? Cause then data is in data.table format - might be confusing.

# Standardise data output

# Data visualisation









```{r hidden, include=FALSE, purl=FALSE}
# Run this at the end to create a R file
options(knitr.duplicate.label = "allow")
knitr::purl('QCPA_r-script.Rmd')
```


